% Abstract v5 - the full post-battery story. Every number here has a script
% and a JSON behind it in the public repo. Nothing enters unstressed.
\begin{abstract}
Load balancing on a graph is usually framed as algorithm design: a better
router yields a better outcome. We present evidence for a two-factor law that
states when the router matters at all. Treating packets as particles under
classical forces, we built independent routing cores---a Newtonian
loss-reaction, an Archimedean buoyancy, a Pascalian pressure diffusion---that
share a search substrate but no mechanism, and used them as instruments on a
bench of fifteen topologies (grids, small-world, scale-free, and the real
G\'EANT and Abilene backbones). Four findings. First, an \emph{attractor}: all
congestion-aware rules produce near-identical per-edge load distributions
(cosine $0.979$, against $0.867$ for blind ECMP; Wilcoxon $p=10^{-4}$ on two
metrics)---the topology and demand fix most of the outcome. Second, an
\emph{interventional law}: scaling link capacities within fixed topologies
shows that congestion \emph{causes} mechanism divergence (rank correlation up
to $1.0$; a squeezed scale-free graph diverges $21\times$, a relieved backbone
converges $8\times$), bounded by a structural ceiling---the real Abilene, our
most congested graph, cannot express divergence its four cycles have no room
for. Third, a \emph{double dissociation}: a structural ratio (tube/sp,
computable before deployment) predicts the gains congestion-aware rerouting
can deliver (partial $r=+0.59$ controlling congestion), while congestion
predicts how much the mechanism's identity matters (partial $r=+0.92$
controlling structure). Fourth, a \emph{validated minimal core}: a two-term
router (gradient plus anticipatory buoyancy) beats the bare substrate on
fifteen of fifteen topologies in every regime ($p\le 10^{-4}$), across a broad
parameter plateau; the reactive scar of our own earlier work does not
replicate, a verdict our re-audit inverted and we report in full. The boundary
is equally explicit: in the saturation regime no physical core reaches
statistical equivalence with the engineered balancers CONGA and DRILL at any
margin---engineering matters exactly where the law says it must.
\end{abstract}
