% Abstract v6 - final. Five findings. Every number has a script and a JSON
% behind it in the public repo. Nothing enters unstressed.
\begin{abstract}
Load balancing on a graph is usually framed as algorithm design: a better
router yields a better outcome. We present evidence for a two-factor law that
states when the router matters at all. Treating packets as particles under
classical forces, we built independent routing cores---a Newtonian
loss-reaction, an Archimedean buoyancy, a Pascalian pressure diffusion, a
Hookean spring---that share a search substrate but no mechanism, and used them
as instruments on a bench of fifteen topologies (grids, small-world,
scale-free, and the real G\'EANT and Abilene backbones). Five findings.
First, an \emph{attractor}: all congestion-aware rules produce near-identical
per-edge load distributions (cosine $0.979$, against $0.867$ for blind ECMP;
Wilcoxon $p=10^{-4}$ on two metrics)---topology and demand fix most of the
outcome. Second, an \emph{interventional law}: scaling link capacities within
fixed topologies shows that congestion \emph{causes} mechanism divergence
(rank correlation up to $1.0$; a squeezed scale-free graph diverges
$21\times$, a relieved backbone converges $8\times$), bounded by a structural
ceiling---the real Abilene, our most congested graph, cannot express
divergence its four cycles leave no room for. Third, a \emph{double
dissociation}: a structural ratio (tube/sp, computable before deployment)
predicts the gains congestion-aware rerouting can deliver (partial $r=+0.59$),
while congestion predicts how much the mechanism's identity matters (partial
$r=+0.92$). Fourth, a \emph{validated minimal core}: gradient plus an
anticipatory threshold force beats the bare substrate on fifteen of fifteen
topologies in every regime ($p\le10^{-4}$), across a broad parameter plateau
and for two functional forms---while the reactive scar of our own earlier
work does not replicate there, a verdict our re-audit inverted and we report
in full. Fifth, a \emph{taxonomy of information channels}: on a
non-stationary bench of episodically failing links, the retired temporal term
returns to beat every memoryless core including the stationary champions
($p\le10^{-3}$), its scar identifying flaky edges with precision
$0.69$--$0.80$ where load statistics reach $0.20$---threshold owns the
nonlinear present, Pascal space, Newton time, and the regime decides which
channel pays. The boundary is equally explicit: in saturation no physical
core reaches statistical equivalence with the engineered balancers CONGA and
DRILL at any margin---engineering matters exactly where the law says it must.
\end{abstract}
