% ====================================================================
% Section: Why the attractor exists - a structural argument.
% The "why", not just the "that": evidence -> explanation.
% Inserted into the attractor section, after the empirical results.
% ====================================================================
\subsection{Why the attractor exists: a structural argument}
\label{sec:attractor-why}

The measurements show \emph{that} independent routers converge to one load
distribution. We now argue \emph{why}, and the argument is structural rather
than physical---which is the point. The load distribution a router produces is
not free; it is constrained by two quantities the router does not control.

\paragraph{The demand fixes the flow across every cut.} Fix a demand matrix
$D$. For any cut $(S, \bar S)$ partitioning the nodes, the total traffic that
must cross from $S$ to $\bar S$ is $\sum_{i \in S, j \in \bar S} D_{ij}$,
\emph{independent of routing}. No choice of paths can change how much flow
traverses a cut; routing only chooses \emph{which} edges of the cut carry it.
By the max-flow/min-cut duality, the narrow cuts---those with few edges or low
capacity relative to the demand crossing them---are forced to carry load up to
their limit regardless of the router. Their utilization is pinned by the
structure, not the algorithm.

\paragraph{Topology sets how many edges are pinned.} In a sparse topology
(a WAN backbone), most demand pairs are separated by narrow cuts: the few
edges on the handful of near-shortest paths are forced to carry nearly all the
traffic, and every reasonable router loads them nearly identically. In a richly
connected topology (a dense grid), most cuts are wide: many edges can share the
load of any cut, leaving the router free to distribute traffic among them in
many near-equivalent ways. The fraction of edges whose load is pinned by the
cut structure is, informally, what the tube/sp ratio measures: low tube/sp
means few alternatives, most edges pinned; high tube/sp means many
alternatives, most edges free.

\paragraph{The attractor is the pinned distribution; the physics fills the
rest.} Putting these together explains every observation in one stroke. The
component of the load distribution fixed by cuts and demand is the attractor:
common to all routers, blind or congestion-aware, because none of them can
move it (this is why even ECMP reaches cosine $0.87$). The residual---the load
on the unpinned edges---is the only part a router can shape, and there any
congestion-aware rule, physical or not, makes the same locally-greedy choice
(spread away from load), which is why the three physical cores converge to
$0.99$ while the blind controls, lacking that rule, do not. Where tube/sp is
low, the pinned component dominates and all routers nearly coincide---the
attractor is strong, and equivalence with engineered routers survives even
strict statistical margins. Where tube/sp is high, the residual is large, the
routers have room to differ, and equivalence becomes margin-sensitive. The
attractor, the convergence of the three physics, the tube/sp predictor, and the
margin-sensitivity of the equivalence are thus four faces of a single
structural fact: \emph{the cut structure of the graph, under a fixed demand,
pins most of the load distribution, and leaves a residual that any
congestion-aware rule shapes identically}.

\paragraph{The argument makes a measurable prediction, and it holds.} The
structural claim is not merely interpretive: it predicts the \emph{size} of the
pinned component as a function of structure. If the attractor is the cut-pinned
floor, then the similarity between a physical core and a congestion-\emph{blind}
router (ECMP)---which can share only that floor, having no rule to shape the
residual---must shrink as tube/sp grows and the residual widens.
It does, sharply: across the fifteen topologies, physics--ECMP similarity falls
with tube/sp at Pearson $r=-0.94$ on cosine and $-0.96$ on the $L_1$ measure
($p<10^{-4}$). The pinned floor ranges from $0.97$ on the most constrained
backbone (Abilene, tube/sp $2.1$) down to $0.77$ on the richest scale-free
graph (tube/sp $9.3$).
The contrast with the physics--physics similarity, which instead \emph{rises}
with tube/sp ($r=+0.54$), is the signature the mechanism predicts: blind
similarity isolates the structural floor and decays as that floor shrinks,
while two congestion-aware cores share the rule that shapes the residual and so
agree \emph{more} precisely exactly where the residual is large. Two routers
blind to congestion measure how much the graph decides; two cores that see it
measure how much the rule decides---the same tube/sp axis governs both, with
opposite sign.

\paragraph{The argument makes a measurable prediction, and it holds.} The
structural claim is not merely interpretive: it predicts the \emph{size} of the
pinned component as a function of structure. If the attractor is the cut-pinned
floor, then the similarity between a physical core and a congestion-\emph{blind}
router (ECMP)---which can share only that floor, having no rule to shape the
residual---must shrink as tube/sp grows and the residual widens.
It does, sharply: across the fifteen topologies, physics--ECMP similarity falls
with tube/sp at Pearson $r=-0.94$ on cosine and $-0.96$ on the $L_1$ measure
($p<10^{-4}$). The pinned floor ranges from $0.97$ on the most constrained
backbone (Abilene, tube/sp $2.1$) down to $0.77$ on the richest scale-free
graph (tube/sp $9.3$).
The contrast with the physics--physics similarity, which instead \emph{rises}
with tube/sp ($r=+0.54$), is the signature the mechanism predicts: blind
similarity isolates the structural floor and decays as that floor shrinks,
while two congestion-aware cores share the rule that shapes the residual and so
agree \emph{more} precisely exactly where the residual is large. Two routers
blind to congestion measure how much the graph decides; two cores that see it
measure how much the rule decides---the same tube/sp axis governs both, with
opposite sign.


\paragraph{A geometric lemma makes the mechanism exact.} The decay has a
clean analytic core. Write each router's normalized utilization as
$u_R = u^\ast + r_R$, where $u^\ast$ is the cut-pinned component every
feasible routing shares and $r_R$ is the residual the router places on the
unpinned edges. Let $f := \lVert u^\ast \rVert^2$ be the pinned fraction of
load energy.
For two routers whose residuals are mutually orthogonal and orthogonal to
$u^\ast$---the defining property of \emph{blind} routers, which place their
residual load without reference to one another---a one-line computation gives
$\cos(u_A, u_B) = \langle u^\ast + r_A,\, u^\ast + r_B\rangle = \lVert u^\ast
\rVert^2 + \langle r_A, r_B\rangle = f$, since both totals are unit-norm and the
residual cross term vanishes. The cosine between two blind routers is, exactly,
the pinned fraction; finite-edge fluctuations are $O(1/\sqrt{E})$ and negligible
on backbone-scale graphs. This is why physics--ECMP similarity tracks $f$: ECMP
spreads load without congestion feedback, so against it the physical core's
agreement is carried entirely by the shared pinned component.
 The lemma also predicts its own boundary: a blind router that
\emph{concentrates} rather than spreads (shortest path) has a residual aligned
with the core's own short-path preference, breaking the orthogonality
assumption---and indeed physics--shortest similarity does not track $f$
(it stays near-flat in tube/sp, $r=+0.46$), while physics--ECMP does
($r=-0.94$). The mechanism is thus not merely consistent with the data; it
specifies which control should reveal it and which should not, and both
predictions hold.


\paragraph{The functional form follows from tube geometry.} The decrease is
not arbitrary either. Of the $w$ edges in the $\alpha$-stretch tube of a
demand pair, only a fraction carry genuine alternative-path flow; the rest are
dead branches the traffic never uses. If a fraction $p$ of the tube absorbs
flow, the pinned share of that pair's load is $sp/(sp + p(w-sp))$, and
with $x = \text{tube/sp} = w/sp$ this is a one-parameter hyperbola
\begin{equation}
f(x) \;=\; \frac{1}{1 + p\,(x-1)}.
\end{equation}
The form is read off the geometry, not fitted post hoc, and it earns its place:
on the fifteen topologies it fits at $R^2=0.89$ and wins on AIC against the
linear and logarithmic alternatives with one fewer parameter, while its single
constant $p=0.037$ is stable to $1.3\%$ under leave-one-out. The shape of
$f(\text{tube/sp})$ is thus pinned down; what stays open is $p$ itself---the
effective fraction of the tube that carries flow---which we measure but do not
yet derive from the graph.

\paragraph{Status of the argument.} Part of this is now proved and part
remains conjectural, and it is worth being exact about which. The geometric
lemma---that the physics--blind cosine equals the pinned fraction $f$---is
established, and its boundary prediction (which control reveals $f$ and which
does not) is confirmed. What is \emph{not} derived in closed form is the
dependence $f(\text{tube/sp})$ itself: we now have its functional form from
tube geometry---a one-parameter hyperbola that wins on AIC---so what stays open
is only the single constant $p$ in it, which we measure ($0.037$, stable to
$1.3\%$) but do not derive from the graph, nor do we prove convergence rates. The four-way coincidence is what the data establish; the lemma
turns it into a mechanism with one proven joint; the closed-form
$f(\text{tube/sp})$ is the conjecture a fuller theory should target.
