% ====================================================================
% Attractor section v3 - corrected numbers (scar-fed Newton, real Abilene,
% 15 genuine topologies), activity conditioning, tube/sp correlation
% handed off to the two-factor section. Supersedes section_attractor_v2.
% ====================================================================
\section{Why They Converge: A Load-Distribution Attractor}
\label{sec:attractor}

Section~\ref{sec:convergence} established \emph{that} the physical cores reach
the same drop-rate performance on most topologies. This section asks
\emph{why}, and the answer turns the convergence from coincidence into
mechanism: the cores do not merely score the same---they route traffic through
the same edges in the same proportions, and so, to a large extent, does every
congestion-aware rule we tested.

\subsection{Measuring the distribution, not the score}

For each of the fifteen topologies we run five routers on an identical graph
and demand: the three physical cores and two blind controls---shortest path
(ignores congestion) and ECMP (hash spread). The Newton core runs with its
scar field alive (an earlier revision of this experiment inadvertently starved
it, degenerating that core to the bare substrate; the bug was caught by our
own audit, fixed, and all numbers below are from the corrected runs). For each
router we record its per-edge utilization vector and compare vectors between
routers with two independent measures: cosine similarity, and one minus the
normalized $L_1$ distance (how much load mass would have to move to turn one
distribution into the other). Twelve seeds per topology.

\subsection{The cores converge to one distribution}

The three physical cores produce near-identical load distributions: cosine
$0.979 \pm 0.009$ and $0.930 \pm 0.018$ on the $L_1$ measure, stable across
the bench. This agreement is not an artifact of shared inactivity: the
activity audit of Section~\ref{sec:gating} shows the cores relocating load by
amounts that differ by an order of magnitude (the scar up to $16\%$ of mass,
the buoyancy at most $3.3\%$), in different directions, on the contested
topologies---and they agree there too (G\'EANT, the most contested graph with
all terms demonstrably active: $0.957$).

\begin{table}[t]
\centering
\caption{Similarity of per-edge load distributions between routers (12 seeds,
15 topologies, corrected runs). ``phys--phys'' is the mean over the three
physical-core pairs; the controls are blind to congestion. Both measures
agree: the cores converge to one distribution, tighter than either control.}
\label{tab:attractor}
\small
\begin{tabular}{lcc}
\toprule
 & cosine & $1-L_1$ \\
\midrule
physics--physics  & $0.979 \pm 0.009$ & $0.930 \pm 0.018$ \\
physics--shortest & $0.938 \pm 0.013$ & $0.837 \pm 0.025$ \\
physics--ECMP     & $0.867 \pm 0.056$ & $0.762 \pm 0.064$ \\
\bottomrule
\end{tabular}
\end{table}

The convergence is significantly tighter than what the blind controls reach: a
paired Wilcoxon test across topologies gives $p = 10^{-4}$ on both measures.
The interpretation is twofold. First, the topology and demand do most of the
work: even ECMP, blind to congestion, lands at cosine $0.87$---the graph
already constrains where load can go. Second, congestion-awareness refines
that rough distribution into a near-consensus ($0.98$) that the blind routers
do not reach, and this shared refinement is what produces the matched drop
rates of Section~\ref{sec:convergence}. The cores tie not by accident but
because they descend into the same basin: a load distribution fixed largely by
the structure and sharpened nearly identically by any congestion-aware rule.

Across topologies, the tightness of this basin correlates with structure
rather than with congestion alone (similarity is highest on the
structure-poor graphs, $r=-0.54$ to $-0.56$ between tube/sp and divergence on
the two measures)---a first sign of the two-factor account that
Section~\ref{sec:twofactor} establishes by intervention.


\medskip
A third, coarser lens agrees. The Jain fairness index of mean per-edge
utilization differs by only $0.016$ across the three physics cores
($0.581$--$0.597$ pooled over the bench), while the engineered balancers
sit visibly apart (DRILL $0.640$; CONGA $0.710$, whose sixteen-path
catalogue spreads load more uniformly by construction). Distinct
mechanics, one basin: catalogue breadth, not mechanism identity, is what
moves the fairness needle.
(\texttt{experiments/jain\_fairness.py},
\texttt{data/jain\_fairness.json})
