% ====================================================================
% v6: the two consensus rocks of review round 2, converted to numbers.
% Stale-state (the crown), per-decision cost, bursty arrivals.
% Every figure here maps to experiments/{stale_state,cpu_bench,
% bursty_bench}.py and data/*.json in the public repository.
% ====================================================================
\subsection{Pricing the deployment assumptions: stale state, decision cost, bursty arrivals}
\label{sec:freshness}

An external review round of nine independent readers converged on exactly the
two limitations declared above: the idealized global state, and the flow-level
granularity. Rather than leave them as confessions, we measured them. Two
predictions were registered before running the staleness experiment: ours, that
degradation at one-twentieth telemetry frequency would stay under half a
percentage point; and a reviewer's, that deciding on stale potential maps would
induce resonance---severe oscillation that amplifies congestion instead of
relieving it. Both turned out to be wrong, in instructive and opposite ways.

\paragraph{Stale state.}
The validated core (gradient plus Archimedean threshold) routes every flow
using a load \emph{view} that is refreshed only every $T$ ticks, while the true
load evolves underneath; $T{=}1$ (one-tick lag, the minimum any real system
pays) is the control. Fifteen topologies, six paired seeds, $T \in
\{1,2,5,10,20,40\}$. Table~\ref{tab:stale} reports the pooled result. The
degradation is real and monotone: negligible at half frequency
($+0.44$pp), material from one-fifth frequency onward ($+2.5$pp at $T{=}5$,
$+7.7$pp at $T{=}20$). Our registered prediction missed the magnitude badly,
and we report that miss in the same spirit as the other inversions in this
paper. The reviewer's predicted failure mode, however, did not appear at all:
the mean temporal variance of per-edge load---the natural signature of
oscillatory rerouting---rises only from $1.00\times$ to $1.29\times$ across
the entire sweep, with no inflection. Stale state makes the core
\emph{blind}, not \emph{unstable}: decisions lose information and quality
degrades smoothly, but the attractor does not turn into a feedback amplifier.
Operationally, the freshness assumption now has a price tag: telemetry at
half the decision rate costs under half a point; at one-fifth the decision
rate the cost consumes the equivalence margin itself.

\begin{table}[t]
\centering
\caption{Stale-state sweep, pooled over 15 topologies $\times$ 6 paired
seeds. The router decides on a load view refreshed every $T$ ticks.
Oscillation ratio is the mean temporal variance of per-edge load relative to
$T{=}1$. (\texttt{experiments/stale\_state.py},
\texttt{data/stale\_state.json})}
\label{tab:stale}
\begin{tabular}{rcccc}
\toprule
$T$ & mean drop & $\Delta$ vs $T{=}1$ (pp) & Wilcoxon $p$ & osc.\ ratio \\
\midrule
1  & 0.028 & ---     & ---                & 1.00 \\
2  & 0.033 & $+0.44$ & $9.9\times10^{-5}$ & 1.02 \\
5  & 0.053 & $+2.49$ & $8.6\times10^{-14}$ & 1.10 \\
10 & 0.082 & $+5.39$ & $8.0\times10^{-16}$ & 1.18 \\
20 & 0.105 & $+7.70$ & $1.2\times10^{-15}$ & 1.26 \\
40 & 0.118 & $+8.94$ & $8.3\times10^{-16}$ & 1.29 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Decision cost.}
On the same substrate (pure Python, identical graphs, warm load state), one
routing decision of the physical core costs 72--2{,}347\,$\mu$s depending on
graph size, the $O(k|E|)$ scaling visible directly. Evaluating CONGA's $K{=}16$
precomputed catalogue costs roughly ten times more \emph{in this substrate},
and DRILL's switch-local sampling is 5--26$\times$ cheaper. None of these
numbers is a hardware measurement---all three mechanisms deploy on optimized
datapaths---but the comparison within a common substrate supports a bounded
claim: the core hides no anomalous decision cost relative to its flow-level
rivals, and at 0.07--2.3\,ms per \emph{flow} (not per packet) it is feasible
in software at backbone flow-arrival rates. Per-packet, microsecond-scale
switching remains DRILL-and-silicon territory.
(\texttt{experiments/cpu\_bench.py})

\paragraph{Bursty arrivals (flow-level).}
A schedule of calm traffic punctuated by dense five-tick bursts of short
flows every twenty-five ticks---the harshest arrival pattern expressible at
flow granularity---opens a gap in DRILL's favour of $+0.59$pp pooled
($p=0.014$): its territory, confirmed and now quantified at roughly half a
point. Thirteen of fifteen topologies remain within the 2pp equivalence
margin. Against CONGA-$K{=}16$ the core concedes nothing (mean $-0.74$pp in
the core's favour, not significant; the mean is driven largely by Abilene,
where the precomputed catalogue pays heavily under bursts, while CONGA wins
ten of fifteen topologies by small margins; fourteen of fifteen within
2pp). True microsecond micro-bursts remain out of scope as stated above.
(\texttt{experiments/bursty\_bench.py}, \texttt{data/bursty\_bench.json})

\paragraph{Partial visibility (the spatial axis).}
The complementary question is spatial: instead of a stale global snapshot,
the deciding node sees the \emph{true} load only on edges within $k$ hops
of itself; beyond that horizon edges show zero load and the potential
degenerates to the static gradient. $k{=}0$ is the blind router; the
blind-to-global gap is $7.7$pp pooled. Two checks were registered before
running: $k{=}2$ recovers at least half the gap, $k{=}3$ at least
two-thirds. Both held (Table~\ref{tab:khop}): one hop of horizon recovers
$36\%$, two hops $63\%$, three hops $77\%$---at which point the core sits
$+1.75$pp from the global view, \emph{inside the equivalence margin
itself}. Vision pays steeply and locally: most of the value of congestion
awareness lives in the deciding node's neighbourhood, where its flows
land first. The physical core does not need a god's-eye view; it needs a
neighbourhood. (\texttt{experiments/khop\_visibility.py},
\texttt{data/khop\_visibility.json})

\begin{table}[t]
\centering
\caption{Spatial visibility sweep, pooled over 15 topologies $\times$ 6
paired seeds. True load is visible within $k$ hops of the deciding node,
zero beyond.}
\label{tab:khop}
\begin{tabular}{rccc}
\toprule
$k$ & mean drop & vs.\ global (pp) & gap recovered \\
\midrule
0 (blind) & 0.105 & $+7.66$ & 0\% \\
1         & 0.077 & $+4.87$ & 36\% \\
2         & 0.056 & $+2.80$ & 63\% \\
3         & 0.046 & $+1.75$ & 77\% \\
global    & 0.028 & ---     & 100\% \\
\bottomrule
\end{tabular}
\end{table}

\medskip
\noindent After this battery, what remains unmeasured: telemetry bandwidth
and propagation dynamics beyond the staleness and visibility knobs above,
and the packet-level regime---a narrower perimeter, still on the queue.
