% ====================================================================
% The ablation, re-audited: the core inverts. Reports the redux (scar
% retired, threshold anticipation crowned), the parameter plateau, and
% the Hooke matched-threshold test that isolates the active ingredient.
% Every number from experiments/{ablation_redux,sensitivity_sweep,
% hooke_quick}.py and the corresponding JSONs in the public repo.
% ====================================================================
\section{The Ablation, Re-Audited: the Core Inverts}
\label{sec:redux}

Section~\ref{sec:ablation} reported the ablation that reduced eight physical
layers to two terms, retaining the Newton-III scar and discarding, among
others, an Archimedean buoyancy. That ablation was run on an earlier harness:
a packet-level simulator, weaker baselines, and---as the audit of
Section~\ref{sec:fair} later showed for the comparison side---conditions that
can flatter or bury a mechanism. Having rebuilt the bench (flow-level
simulation, fifteen genuine topologies including the real Abilene, paired
seeds), we re-ran the foundational question: does each physical term improve
drop rate over the bare substrate (gradient plus linear congestion), and
where?

\subsection{The verdict inverts}

It does not survive. The scar---the term the original ablation kept---yields
no significant improvement in its home regime (contested topologies:
$-0.34$\,pp, $p=0.12$), \emph{worsens} the dense grids, and degrades the
saturated backbone by $+2.1$\,pp: a reactive memory of failures is largely
redundant with the congestion term that already steers away from load, and in
saturation it over-scars and thrashes. It is also the term that relocates the
most load mass ($6$--$16\%$, Section~\ref{sec:gating})---movement without
payoff. The buoyancy---the term the original ablation discarded---beats the
substrate in every regime (free flow $-0.24$\,pp, $p=10^{-4}$; contested
$-1.05$\,pp, $p<10^{-5}$; saturated $-3.31$\,pp, $p=5\times10^{-4}$) and on
fifteen of fifteen topologies without exception (sign test
$p=6\times10^{-5}$). The original verdict was a harness artifact, of the same
family as the path-budget artifact of Section~\ref{sec:fair}; we report the
inversion rather than quietly rewriting history.

\subsection{Not a needle: the parameter plateau}

A natural objection is overfitting: perhaps the buoyancy wins only at its
default parameters. We swept the flotation threshold ($\rho_0$ from $0.4$ to
$0.8$), the gain ($g$ from $2$ to $32$), and the substrate's congestion weight
($\beta$ from $1.5$ to $6$). In all fourteen settings the buoyancy beats the
substrate on fifteen of fifteen topologies ($p$ between $10^{-4}$ and
$10^{-12}$). The trends are physically sensible: earlier anticipation (lower
$\rho_0$) and stronger response help more, and a weaker substrate leaves more
for the buoyancy to do. The win is a plateau, not a needle.

\subsection{What the ingredient is: threshold, not curvature}

The buoyancy differs from the substrate's linear congestion term in two ways
at once: a threshold (no force below $\rho_0$) and a curvature (quadratic
growth above it). A fourth classical core separates them: a Hookean spring,
linear above a slack threshold. The spring also beats the substrate on fifteen
of fifteen topologies (contested $-1.50$\,pp, $p<10^{-13}$; saturated
$-4.38$\,pp), and at a matched threshold the linear and quadratic forms
deliver nearly identical pooled gains ($-1.2$ versus $-1.15$\,pp). The active
ingredient is therefore the \emph{anticipatory threshold}---a force that
engages before loss, above a density the linear substrate treats as
ordinary---while the functional form above the threshold is secondary. The
spring also lands in the attractor with its siblings (cosine $0.987$ to the
other cores), a fourth witness to Section~\ref{sec:attractor}.

\subsection{What ``the core'' now means}

The validated two-term core of this paper is \emph{gradient plus anticipatory
threshold force}: implement the threshold with a buoyancy or a spring and the
result stands, across the parameter plateau, in every congestion regime, on
every topology of the bench. The reactive scar that founded this project is
retired from the core with the role it earned: the instrument that started the
search, and one of the convergence witnesses---but not, on the modern bench, a
source of routing value. We consider catching this inversion ourselves, with
the same audit machinery that earlier caught an inflated advantage over CONGA,
to be part of the contribution.
