% ====================================================================
% The temporal channel: Newton re-examined, and the taxonomy that
% closes the circle. Every number from experiments/newton_redemption.py
% and data/newton_redemption.json. The flaky bench is declared as
% designed a posteriori for memory habitat: existence claim, not
% generality.
% ====================================================================
\section{The Temporal Channel: a Taxonomy of Physical Information}
\label{sec:taxonomy}

Section~\ref{sec:redux} retired the Newton-III scar from the stationary core,
and left a question a careful reader should ask: was the law wrong, or was the
bench wrong for the law? Two observations argue for a second look. First,
calibration: the scar parameters were never swept, and the historical decay of
$0.998$ per tick gives the scar a half-life of $346$ ticks in a $200$-tick
simulation---the ``reaction'' of the redux never healed; it was a permanent
grudge, and permanent over-punishment is a sufficient explanation for the
thrashing observed on dense grids and in saturation. Second, and more
fundamental, habitat: the scar is \emph{temporal memory}, and the entire bench
of this paper is \emph{stationary}---fixed demand statistics, fixed
capacities. In a stationary world the instantaneous state is statistically
sufficient and memory is redundant \emph{by construction of the bench}, not by
fault of the law. We had evaluated a historian in a world without a past.

\subsection{Calibration accounts for the harm, not for a stationary win}

Re-running the stationary bench with swept scar parameters confirms both
halves. Short half-lives remove the damage (the saturated backbone flips from
$+2.1$\,pp \emph{worse} to slightly better than the substrate; the dense-grid
harm disappears at gentle deposits), so the redux's negative verdict was in
part an artifact of temporal miscalibration. But no calibration produces a
clear stationary win either: the verdict of Section~\ref{sec:redux}
stands---where there is no history, memory has nothing to sell.

\subsection{A bench with a past: flaky links}

The falsifiable question is therefore whether a realistic regime exists in
which temporal memory is the winning channel. Real networks supply an obvious
candidate: \emph{flaky links}---hardware that fails in episodes and looks
healthy in between, precisely the case where the instantaneous state misleads
and only memory can warn. We built a non-stationary bench: in each run,
$15\%$ of edges (fixed per seed, shared by all routers) suffer periodic
degradation episodes (every $40$ ticks, for $8$ ticks, capacity scaled to
$0.2$), recovering fully in between. We state plainly that this bench was
designed \emph{a posteriori} for memory's habitat: the claim defended here is
one of existence---there is a realistic regime where the temporal channel
wins---not of generality.

In that regime the retired physicist wins outright. The calibrated scar
(half-life matched to the episode period) beats the bare substrate on
fourteen of fifteen topologies ($-1.26$\,pp, $p=4\times10^{-10}$), and---the
decisive comparison---beats both threshold cores as well: Archimedes on
thirteen of fifteen ($p=5\times10^{-4}$) and Hooke on twelve of fifteen
($p=9\times10^{-3}$). Its sole loss is the saturated backbone, exactly where
the two-factor law of Section~\ref{sec:twofactor} says no mechanism can act.
An anti-ghost check confirms the channel is real: ranking edges by
accumulated scar identifies the flaky set with precision $0.69$--$0.80$,
while ranking by mean utilization---the best a memoryless router could
do---reaches only $0.06$--$0.20$. The scar sees what load cannot. A
calibration law emerges for free: the long-memory (historical) scar performs
even better when the failing set is persistent ($-1.68$\,pp pooled) and
drowns in saturation---the scar's half-life should match the timescale of
the disturbance it is meant to remember.

\subsection{The taxonomy}

The arc of this paper's cores now closes into a taxonomy of information
channels. The substrate reads the \emph{linear present}. The threshold cores
(Archimedes, Hooke) add the \emph{nonlinear present}---anticipatory warning
above a density the linear term treats as ordinary---and own the stationary
regime (Section~\ref{sec:redux}). Pascal adds \emph{space}---neighbourhood
pressure---and contributes modestly wherever contention is local. Newton adds
\emph{time}---memory of failures---and owns the non-stationary regime, where
it beats every memoryless core including the stationary champions. No channel
dominates everywhere; the regime decides which information pays, in the sense
made quantitative by the two-factor law. The practical reading for a network
operator is correspondingly plain: a two-term gradient-plus-threshold rule
suffices wherever traffic is stationary; add the temporal term, with its
half-life matched to failure timescales, on infrastructure known to flap.
